Is Vagueness Beneficial for Hearers? \protect \\ Evidence from Experiments

### 2016-09-21 papers/2016-Special-Issue/v-special-issue-2016.tex
Kees's email said: The changes you’ve made look to me to be on the right track. Just in case you’ve got time to look at this again before tomorrow, I’ve revised the closing paragraphs of the paper a bit (where applications to NLG are discussed).

Language use may be called vague for various reasons.\footnote{See e.g. the entry ``vague'' in \cite{penguin}.} In most academic use though, the word `vagueness' has a specific meaning. Keefe and Smith, for example, state ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \cite[p.\ 4]{keefe1996vagueness}, also \citeA{EgreKlinedinst}).  The crucial criterion is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \cite[p.\ 1]{lipmanvague}. A typical example is the word ``tall'', as applied to people for example, because here is no precise, known height which separates those who are tall from those who are not. The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

@@

One group of participants was given HHI scores in the ideal range: for this group their weight loss did not differ depending on whether they were given vague or precise HHI values. However for the other group, who were given HHI scores outside the ideal range, their weight loss was significantly greater if they were given vague HHI scores than if they were given precise HHI scores. The authors explain the improvement in the vague condition for this group as resulting from the participants' freedom to think of themselves a positioned on one end of the range - the end closest to the ideal HHI scores. This ``illusion of proximity'' \cite<>[p.~4]{Mishra01042011} to the goal is argued to allow participants to generate positive expectancies that lead to behaviours that improve performance. In contrast, in the precise conditions, participants did not have this freedom of interpretation, and could not distort the information to bring about the beneficial \emph{illusion of proximity}. These results are interesting, and of obvious potential practical importance. We note, however, that information presented as an exact range of values does not conform with the standard definition of vagueness \cite{keefe1996vagueness, EgreKlinedinst}, since an exact range does not admit borderline cases. In the terminology of \citeA{Hobbs85granularity}, the difference between a range and a single midpoint value is a difference of \emph{granularity}. Furthermore, the experiments of \citeA{Mishra01042011} did not explore benefits in terms of processing cost, but in terms of long-term behaviour change.

@@

A comparison may clarify the logic of the situation. In recent years a number of studies, focussing typically on red wine, have suggested that alcohol, consumed in low doses, may have health benefits. An alternative explanation, however, asserts that it is not the alcohol in the wine that was beneficial, but antioxidants from grapes. If this alternative explanation is correct then alcohol may not be as beneficial as some would like to think.\\[2ex]
%
{\bf Implications for practical NLG systems.} Our findings suggest a re-think of the questions on which much research on the utility of vagueness rests. The question of how a particular piece of quantitative information is best conveyed through language is certainly an important one for practical NLG. Years of research on the logic of vagueness in natural language have primed the research community to expect that the utility of vagueness is an important part of the answer, but our findings call this into question. %It would appear that this is true for most vague expressions, and not only the ones investigated in our experiments, where a speaker is referring to a square with dots.

Consider, once again, a number of ways in which a given temperature can be reported. Given a temperature measurement (or prediction) as input, an NLG system might justifiably say that it was
%
\begin{quote}
(a) {\em 27.2 degrees}, or\\
(b) {\em approximately 27 degrees}, or\\
(c) {\em above 25 degrees}, or\\
(d) {\em warm}, 
\end{quote}
%
among other candidate expressions. If standard accounts are to be believed, then options (a) and (c) convey crisp information, whereas (b) and (d) are vague (i.e., they permit borderline cases). But, to the best of our knowledge, there is no experimental evidence that the fact in itself that an expression is vague confers a benefit (or a de-benefit) on it for hearers. Rather than asking whether a candidate expression is vague, other questions might shed more light on the NLG system's choice, similar (though not necessarily identical) to the ones identified in our empirical studies. For a start, both (c) and (d) designate an interval that is half-bounded, whereas both (a) and (b) express closed intervals, and this might affect their comprehension by a hearer (cf., our finding about comparison versus matching strategies). Furthermore, the expressions (c) and (d) -- one of which is vague while the other is crisp -- appear to convey very similar amounts of information, saying that the temperature is higher than some (fairly high) standard. Perhaps most importantly, "warm" is shorter than the other three expressions and avoids the use of numbers, and our experiments suggest that this may make ``warm" more rapidly understood than its competitors; earlier experiments point in the same direction, given the evaluative nature of the word ``warm" (see section 1 and \citeA{peters2009bringing}).

One way to see why vagueness may not matter as much for NLG -- and for human communication more generally -- as is often thought is the following thought experiment. Many speakers understand the word ``warm" as vague. Now suppose a group of users agreed to give it a precise definition; according to this precisification, ``warm" means $>25$ degrees (as in (c)). It seems unlikely that this change, from a vague meaning to a crisp one, would change the utility of the word.

>>>>

Language use may be called vague for various reasons.\footnote{See e.g. the entry ``vague'' in \cite{penguin}.} In most academic use though, the word `vagueness' has a specific meaning. Keefe and Smith, for example, state ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \cite[p.\ 4]{keefe1997vagueness}, also \citeA{EgreKlinedinst}).  The crucial criterion is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \cite[p.\ 1]{lipmanvague}. A typical example is the word ``tall'', as applied to people for example, because here is no precise, known height which separates those who are tall from those who are not. The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

@@

One group of participants was given HHI scores in the ideal range: for this group their weight loss did not differ depending on whether they were given vague or precise HHI values. However for the other group, who were given HHI scores outside the ideal range, their weight loss was significantly greater if they were given vague HHI scores than if they were given precise HHI scores. The authors explain the improvement in the vague condition for this group as resulting from the participants' freedom to think of themselves a positioned on one end of the range - the end closest to the ideal HHI scores. This ``illusion of proximity'' \cite<>[p.~4]{Mishra01042011} to the goal is argued to allow participants to generate positive expectancies that lead to behaviours that improve performance. In contrast, in the precise conditions, participants did not have this freedom of interpretation, and could not distort the information to bring about the beneficial \emph{illusion of proximity}. These results are interesting, and of obvious potential practical importance. We note, however, that information presented as an exact range of values does not conform with the standard definition of vagueness \cite{keefe1997vagueness, EgreKlinedinst}, since an exact range does not admit borderline cases. In the terminology of \citeA{Hobbs85granularity}, the difference between a range and a single midpoint value is a difference of \emph{granularity}. Furthermore, the experiments of \citeA{Mishra01042011} did not explore benefits in terms of processing cost, but in terms of long-term behaviour change.

@@

A comparison may clarify the logic of the situation. In recent years a number of studies, focussing typically on red wine, have suggested that alcohol, consumed in low doses, may have health benefits. An alternative explanation, however, asserts that it is not the alcohol in the wine that was beneficial, but antioxidants from grapes. If this alternative explanation is correct then alcohol may not be as beneficial as some would like to think.\\[2ex]
%
{\bf Implications for practical NLG systems.} Our findings suggest a re-think of the questions on which much research on the utility of vagueness rests. The question of how a particular piece of quantitative information is best conveyed through language is an important one for practical NLG. Years of research on the logic of vagueness in natural language have primed the research community to expect that the utility of vagueness is an important part of the answer, but our findings call this expectation into question. Although our own studies have focussed on vagueness in descriptive Noun Phrases, it seems plausible that vagueness plays a similar role in other linguistic constructs.

For example, consider air temperature, once again. Given a temperature measurement (or prediction) as input, an NLG system that addresses a non-expert audience might say that it was (or will be)
%
\begin{quote}
(a) {\em 27.2 degrees}, or\\
(b) {\em approximately 27 degrees}, or\\
(c) {\em above 25 degrees}, or\\
(d) {\em warm}, 
\end{quote}
%
among other candidate expressions. If standard accounts are to be believed, then options (a) and (c) convey crisp information, whereas (b) and (d) are vague (i.e., they permit borderline cases). From our own experiments or the ones we found in the literature, there is no evidence that the fact in itself that an expression is vague should confer a benefit on it for hearers. Rather than asking whether a candidate expression is vague, other questions might shed more light on the NLG system's choice, similar (though not necessarily identical) to the ones identified in our studies. These questions might focus on the amount of information that a given expression conveys (i.e., on granularity), on the avoidance of numbers, and on the use of evaluative terms. Let's see how this might pan out for the examples at hand.

The expressions (c) and (d) -- one of which is vague while the other is crisp -- convey similar amounts of information, saying that the temperature is higher than some standard (i.e., designating a half-open temperature interval, as opposed to the closed intervals designated by (a) and (b)). Yet (d) (``warm") is shorter than the other three expressions and avoids the use of numbers, and our experiments suggest that this may make ``warm" more rapidly understood than its competitors; earlier experiments point in the same direction, given the evaluative nature of ``warm" (see section 1). Results of this kind can be applied in NLG, though obvious care would need to be exercised to take the purpose of the communication into account.\footnote{In the experiments reported in this paper, the purpose of the utterances investigated was always clear (e.g., instructing people on a weight loss program \cite{Mishra01042011}; informing a patient's choice of hospital \cite{peters2009bringing}; or selecting a dot array (in our own studies).}

One way to see why vagueness may not matter as much for NLG -- and for human communication -- as is often thought, is the following thought experiment. Suppose a group of speakers understand the word ``warm" as vague, agreeing that temperatures above 26 count as warm, and temperatures below 24 do not count as warm, but considering temperatures between 24 and 26 as borderline cases. Now suppose these speakers agreed to ``precisify" their definition deciding that, henceforth, ``warm" means ``$>25$ degrees" (as in (c) above): this decision resolves the borderline cases, while everything else remains the same. It seems unlikely that this change in language use, from a vague meaning to a crisp one (i.e., one that has no borderline cases), would lower the utility of the word. Our experimental findings are consistent with this idea.

%\bibliography{bibliography}
